A Mechanised Library for Generic (Co)Programming

### 2024-02-27 paper/paper.tex
Stuff

\subsection{Types and Morphisms}
\label{sec:types-morphisms}
We address the different forms of equality by defining a class of setoids,
i.e.\ types with na associated equivalence relation, and considering only
functions that respect the associated equivalences, i.e. \emph{proper}
morphisms with respect to the function \emph{respectfulness} relation.  We use
the type-class mechanism, instead of setoids in Coq's standard library, to help
Coq's code extraction mechanism remove any occurence of custom equivalence
relations in the extracted OCaml code.  We use \coq{=e} to denote
the equivalence relation of a setoid. By default, we associate every Coq type
with the standard propositional equality, unless a different equivalence is
specified (i.e.\ we allow overlapping instances, and Coq's propositional
equality takes the lowest priority). 
\begin{coqcode}
Reserved Notation "f =e g" (at level 70, no associativity).
Class setoid A : Type :=
  MkEquiv
    { eqRel : A -> A -> Prop;
      e_refl : forall x, eqRel x x;
      e_sym : forall x y, eqRel x y -> eqRel y x;
      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
    }.
Notation "f =e g" := (eqRel f g).
\end{coqcode}
Given types \coq{A} and  \coq{B}, with their respective
equivalence relations \coq{eA : A -> A -> Prop} and
\coq{eB : B -> B -> Prop}, the we define the type
\coq{A ~> B} to represents \emph{proper} morphisms of the
respectfulness relation of \coq{R} to \coq{R'}.
\begin{coqcode}
Structure morph A {eA : setoid A} B {eB : setoid B} :=
  MkMorph
    { app :> A -> B;
      app_eq : forall x y, x =e y -> app x =e app y
    }.
Notation "A ~> B" := (@morph A _ B _).
\end{coqcode}
Note that we leverage on Coq's type class mechanism to fill in the necessary
equivalence relations. Since Coq's code extraction mechanism will erase any
occurrence of \coq{Prop} in the code, so objects of type
\coq{A ~> B} will be extracted to the OCaml equivalent to
\coq{A -> B}.

@@

\subsubsection{Isomorphisms of Coq Types and Container Extensions}

Using custom equivalences and proper morphisms, we redefine the definitions of
container extensions and container equality. In particular, container
extensions require a proper morphism to check the validity of positions in
shapes, and container equality now uses equivalences of shapes and contained
elements: 
\begin{coqcode}
Class Cont (Shape : Type) (Pos : Type) := { valid : Shape ~> Pos ~> bool }.
Inductive AppR `{e : setoid X} (x y : App F X) : Prop :=
  | AppR_ext (Es : shape x =e shape y)
             (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
\end{coqcode}
Note the use of \coq{=e} instead of Coq's standard equality.  For
positions, however, we chose to use Coq's propositional equality, since these
would leads to simpler code. We explain why in Section~\ref{sec:cata}, and the
mechanisation of initial algebras of container extensions. The definitions in
Figure~\ref{fig:poly} are redefined to account for these changes in a
straightforward way.

We prove the isomorphisms of polynomial types and the equivalent (redefined)
container extensions. As an example, we discuss (informally) the isomorphisms
of pairs with their equivalent container extensions. Suppose that we
know that \coq{App F X} is isomorphic to \coq{A}, and
\coq{App G X} is isomorphic to \coq{B}. Then we can
show that \coq{App (Prod F G) X} is isomorphic to
\coq{A * B}.  If we have an element of type 
\coq{App (Prod F G) X},
using the \coq{inl} position, we can obtain
\coq{App F X}.  Similarly, using \coq{inr}, we can
obtain \coq{App G X}. Since these are the only two valid positions
in the shape of pairs, we have finished. It is now sufficient to use the
isomorphisms of \coq{App F X} and \coq{App G X} to
obtain \coq{A * B}. Similarly if we have \coq{A * B},
we can first use the isomorphisms of 
\coq{A} and
\coq{B} to obtain
\coq{App F X * App G X}, and then construct the necessary container
extension. Given \coq{p_inl}/\coq{p_inr} that act as
\coq{inl/inr} but on valid product positions, the relevant
functions that define the isomorphism are:
\begin{coqcode}
Definition iso_pair (x : App (Prod F G) X) : App F X * App G X :=
  ({| shape := shape (fst x); cont := fun e => cont x (e_inl e) |}, 
   {| shape := shape (snd x); cont := fun e => cont x (e_inr e) |}).
Definition iso_prod (x : App F X * App G X) : App (Prod F G) X :=
{| shape := (shape (fst x), shape (snd x));
   cont := fun e => match val e with 
                    | inl p => fun H => cont (fst x) (ValidPos p H)
                    | inr p => fun H => cont (snd x) (ValidPos p H)
                    end (IsValid e)
|}.
\end{coqcode}
Proving that these functions indeed form an isomorphism is straightforward,
using the fact that \coq{Prod} containers only have two valid positions.

\subsubsection{Generalised Rewriting for Morphisms}

Since our framework only allows types with equivalences, and proper morphisms,
we can leverage Coq's \emph{setoid rewrite}

\dcas{Blabla: examples, no need to explain setoid rewrite in detail here}

>>>>

\subsection{Types and Morphisms}
\label{sec:types-morphisms}
We address the different forms of equality by defining a class of setoids,
i.e.\ types with na associated equivalence relation, and considering only
functions that respect the associated equivalences, i.e. \emph{proper}
morphisms with respect to the function \emph{respectfulness} relation.  We use
the type-class mechanism, instead of setoids in Coq's standard library, to help
Coq's code extraction mechanism remove any occurence of custom equivalence
relations in the extracted OCaml code.  We use \coq{=e} to denote
the equivalence relation of a setoid. By default, we associate every Coq type
with the standard propositional equality, unless a different equivalence is
specified (i.e.\ we allow overlapping instances, and Coq's propositional
equality takes the lowest priority). 
\begin{coqcode}
Reserved Notation "f =e g" (at level 70, no associativity).
Class setoid A : Type :=
  MkEquiv
    { eqRel : A -> A -> Prop;
      e_refl : forall x, eqRel x x;
      e_sym : forall x y, eqRel x y -> eqRel y x;
      e_trans : forall x y z, eqRel x y -> eqRel y z -> eqRel x z;
    }.
Notation "f =e g" := (eqRel f g).
\end{coqcode}
Given types \coq{A} and  \coq{B}, with their respective
equivalence relations \coq{eA : A -> A -> Prop} and
\coq{eB : B -> B -> Prop}, the we define the type
\coq{A ~> B} to represents \emph{proper} morphisms of the
respectfulness relation of \coq{R} to \coq{R'}.
\begin{coqcode}
Structure morph A {eA : setoid A} B {eB : setoid B} :=
  MkMorph
    { app :> A -> B;
      app_eq : forall x y, x =e y -> app x =e app y
    }.
Notation "A ~> B" := (@morph A _ B _).
\end{coqcode}
Note that we leverage on Coq's type class mechanism to fill in the necessary
equivalence relations. Since Coq's code extraction mechanism will erase any
occurrence of \coq{Prop} in the code, so objects of type
\coq{A ~> B} will be extracted to the OCaml equivalent to
\coq{A -> B}. Furthermore, we add a coercion from \coq{A ~> B} to \coq{A -> B}.
On top of this, we define basic function composition and identity functions:
\begin{coqcode}
Notation "f \o g" = (comp f g).
Definition comp : (B ~> C) ~> (A ~> B) ~> A ~> C := ...
Definition id : A ~> A := ...
\end{coqcode}

Using custom equivalences and proper morphisms, we redefine the definitions of
container extensions and container equality. In particular, container
extensions require a proper morphism to check the validity of positions in
shapes, and container equality now uses equivalences of shapes and contained
elements: 
\begin{coqcode}
Class Cont (Shape : Type) (Pos : Type) := { valid : Shape * Pos ~> bool }.
Inductive AppR `{e : setoid X} (x y : App F X) : Prop :=
  | AppR_ext (Es : shape x =e shape y)
             (Ek : forall e1 e2, val e1 = val e2 -> cont x e1 =e cont y e2).
\end{coqcode}
Note the use of \coq{=e} instead of Coq's standard equality.  For positions,
however, we chose to use Coq's propositional equality, since these would leads
to simpler code. We explain why in Section~\ref{sec:cata}, and the
mechanisation of initial algebras of container extensions. The definitions in
Figure~\ref{fig:poly} are redefined to account for these changes in a
straightforward way, i.e. we prove that validity predicates are indeed proper
morphisms.  \dcas{Should we show, instead, the actual full definition? It
becomes a bit more complicated ...}

This definition leads to the well-known ``setoid hell'', which we mitigate by
providing tactics and notations to automatically discharge proofs of
\coq{app_eq} for morphisms, whenever the types use the standard propositional
equality, or a combination of propositional and extensional equality. However,
our compositional approach allows to build morphisms by plugging in other
morphisms to our combinator, In our framework, our goal is that the
user-provided functions remain small, with relatively straightforward proofs of
\coq{app_eq}.

However, by using this mechanisation, we gain simplified proofs via the use of
Coq's \emph{Generalised Rewriting}.  Since every morphism \coq{f : A ~> B}
satisfies that if \coq{x =e y}, then \coq{f x =e f y}, we can add every
morphism as a proper element of Coq's respectfulness relation. In practice,
this means that we can use the \coq{rewrite} tactic on proofs of type 
\coq{A =e B}, for arbitrary \coq{A} and \coq{B}, whenever they are used as
arguments of morphisms, as well as Coq's \coq{reflexivity}, \coq{symmmetry},
and \coq{transitivity} tactics. For example:
\begin{coqcode}
Goal forall `(f : A ~> B) `(g : B ~> C) `(h : C ~> D), 
  h \o g =e id -> h \o (g \o f) =e f.
Proof.
  intros H.  rewrite compA H.  reflexivity.
Qed.
\end{coqcode}

@@

We prove the isomorphisms of polynomial types and the equivalent (redefined)
container extensions. As an example, we discuss (informally) the isomorphisms
of pairs with their equivalent container extensions. Suppose that we
know that \coq{App F X} is isomorphic to \coq{A}, and
\coq{App G X} is isomorphic to \coq{B}. Then we can
show that \coq{App (Prod F G) X} is isomorphic to
\coq{A * B}.  If we have an element of type 
\coq{App (Prod F G) X},
using the \coq{inl} position, we can obtain
\coq{App F X}.  Similarly, using \coq{inr}, we can
obtain \coq{App G X}. Since these are the only two valid positions
in the shape of pairs, we have finished. It is now sufficient to use the
isomorphisms of \coq{App F X} and \coq{App G X} to
obtain \coq{A * B}. Similarly if we have \coq{A * B},
we can first use the isomorphisms of 
\coq{A} and
\coq{B} to obtain
\coq{App F X * App G X}, and then construct the necessary container
extension. Given \coq{p_inl}/\coq{p_inr} that act as
\coq{inl/inr} on valid product positions, the relevant
functions that define the isomorphism are:
\begin{coqcode}
Definition iso_pair (x : App (Prod F G) X) : App F X * App G X :=
  ({| shape := shape (fst x); cont := fun e => cont x (e_inl e) |}, 
   {| shape := shape (snd x); cont := fun e => cont x (e_inr e) |}).
Definition iso_prod (x : App F X * App G X) : App (Prod F G) X :=
{| shape := (shape (fst x), shape (snd x));
   cont := fun e => match val e with 
                    | inl p => fun H => cont (fst x) (ValidPos p H)
                    | inr p => fun H => cont (snd x) (ValidPos p H)
                    end (IsValid e)
|}.
\end{coqcode}
Proving that these functions indeed form an isomorphism is straightforward,
using the fact that \coq{Prod} containers only have two valid positions.
